Rapid-response Bayesian sky localization for electromagnetic follow-up of gravitational-wave candidates

### 2013-02-12 ms.tex
Remove nitty-gritty Fisher matrix derivation
Removed double-space separating sentences
Tightened up Fisher matrix derivation

When a \ac{CBC} occurs, a detector's strain observation may be described by a linear combination of two orthogonal \ac{GW} quadratures and additive, zero\nobreakdashes-mean, \acl{WSS}, Gaussian measurement noise $n(t)$, with one-sided power spectral density $S(f)$. In the frequency domain, the detector's observation $\tilde x (\omega)$ is given by
%
\begin{equation}\label{eq:signal-model}
	\tilde x (\omega) = \tilde h (\omega, \rho, \gamma, \tau) + \tilde n (\omega) = \rho e^{\imath (\gamma - \omega \tau)} \tilde u (\omega) + \tilde n (\omega).
\end{equation}
%
where $\tilde h (\omega)$ is the signal content, $\tilde n (\omega)$ is the detector's noise, $\tilde u (\omega)$ is the frequency-domain post-Newtonian inspiral signal, $\rho$ is the signal's amplitude, $\gamma$ is a phase angle the encodes the relative strength of the two \ac{GW} quadratures, and $\tau$ is the time of arrival relative to an arbitrary time base.

The Fisher information matrix for a measurement $\tilde x$ described by the unknown parameter vector $\boldsymbol{\theta}$ may be written as
%
\begin{equation}\label{eq:general-fisher-matrix}
	\mathcal{I}_{jk} = \mathrm{E} \, \left[
		-\frac{\partial^2 \log
			\mathcal{L}(\tilde x ; \boldsymbol\theta)}
			{\partial \theta_j \theta_k}
	\right].
\end{equation}
%
This equation involves double derivatives, but substantial simplification is possible when the likelihood is Gaussian. First, we write the likelihood of the data $\tilde x (\omega)$ given the parameter vector $\boldsymbol\theta = (\rho, \gamma, \tau)$:
%
\begin{equation}\label{eq:gaussian-likelihood}
	\mathcal{L}(\tilde x; \boldsymbol\theta) = p(\tilde x | \boldsymbol\theta)
		\propto \exp \left[
		- \frac{1}{2} \int_0^\infty \frac{\left|\tilde x (\omega)
			- \tilde h (\omega; \boldsymbol\theta) \right|^2}{S(\omega)} \, \mathrm{d}\omega
	\right].
\end{equation}
%
The normalization has been dropped because the Fisher matrix elements depend only on derivatives of the log\nobreakdashes-likelihood function and any factor that is independent of the parameters makes no contribution. Now, substituting \eref{eq:gaussian-likelihood} into \eref{eq:general-fisher-matrix} gives us
%
\begin{equation}\label{eq:gaussian-fisher-matrix}
	\mathcal{I}_{jk} = \int_0^\infty \Re \left[
        \left(\frac{\partial \tilde h}{\partial \theta_j}\right)^*
        \left(\frac{\partial \tilde h}{\partial \theta_k}\right)
	\right] \frac{1}{S(\omega)} \, \mathrm{d}\omega.
\end{equation}
%
where $\Re$ denotes the real part and $^*$ the complex conjugate. This form is more useful because it involves only first derivatives, and of the signal $\tilde h (\omega)$ rather than the entire observation $\tilde x (\omega)$.

With \eref{eq:gaussian-fisher-matrix} in hand, we return to our particular signal model, \Eref{eq:signal-model}. The necessary partial derivatives are
%
\begin{eqnarray*}
    \frac{\partial \tilde h}{\partial \rho} &= e^{\imath (\gamma - \omega \tau)} \tilde u (\omega) \\
    \frac{\partial \tilde h}{\partial \gamma} &= \imath \, \rho \, e^{\imath (\gamma - \omega \tau)} \tilde u (\omega) \\
    \frac{\partial \tilde h}{\partial \tau} &= -\imath \, \omega \, \rho \, e^{\imath (\gamma - \omega \tau)} \tilde u (\omega).
\end{eqnarray*}
%
The Fisher matrix elements are
%
\begin{eqnarray*}
    \mathcal{I}_{\rho\rho} &= \int_0^\infty \frac{|\tilde u (\omega)|^2}{S(\omega)} \, \mathrm{d}\omega \\
	\mathcal{I}_{\gamma\gamma} &= \rho^2 \int_0^\infty \frac{|\tilde u (\omega)|^2}{S(\omega)} \, \mathrm{d}\omega \\
	\mathcal{I}_{\tau\tau} &= \rho^2 \int_0^\infty \frac{|\tilde u (\omega)|^2}{S(\omega)} \omega^2 \, \mathrm{d}\omega \\
	\mathcal{I}_{\gamma\tau} &= -\rho^2 \int_0^\infty \frac{|\tilde u (\omega)|^2}{S(\omega)} \omega \, \mathrm{d}\omega = \mathcal{I}_{\tau\gamma} \\
	\mathcal{I}_{\rho\gamma} &= \mathcal{I}_{\gamma\rho} = \mathcal{I}_{\rho\tau} = \mathcal{I}_{\tau\rho} = 0.
\end{eqnarray*}
%
If the waveform $\tilde u (\omega)$ is normalized such that $\mathcal{I}_{\rho\rho} = 1$, we can introduce the \textit{\ac{SNR}-weighted moments of angular frequency},
%
\begin{equation}\label{eq:angular-frequency-moments}
    \overline{\omega^k} =
        \left[ \int_0^\infty \frac{|\tilde u (\omega)|^2}{S(\omega)} \omega^k \, \mathrm{d}\omega \right]
        \left[ \int_0^\infty \frac{|\tilde u (\omega)|^2}{S(\omega)} \, \mathrm{d}\omega \right]
\end{equation}
%
Then, we have
%
\begin{eqnarray*}
	\mathcal{I} = \bordermatrix{
        ~ & \rho & \gamma & \tau \cr
        \rho & 1 & 0 & 0 \cr
        \gamma & 0 & \rho^2 & -\rho^2 \overline{\omega} \cr
        \tau & 0 & -\rho^2 \overline{\omega} & \rho^2 \overline{\omega^2}
    }.
\end{eqnarray*}

>>>>

In the frequency domain, the strain $\tilde x (\omega)$ observed by a detector is a linear combination of two orthogonal \ac{GW} quadratures and additive noise,
%
\begin{equation}\label{eq:signal-model}
	\tilde x (\omega) = \tilde h (\omega, \rho, \gamma, \tau) + \tilde n (\omega) = \rho e^{\imath (\gamma - \omega \tau)} \tilde u (\omega) + \tilde n (\omega),
\end{equation}
%
where $\tilde h (\omega)$ is the signal content, $\tilde n (\omega)$ is the detector's \ac{WSS} Gaussian noise with \ac{PSD} $S(\omega)$, $\tilde u (\omega)$ is the frequency-domain post-Newtonian inspiral signal, $\rho$ is the signal's amplitude, $\gamma$ is a phase angle the encodes the relative strength of two \ac{GW} quadratures, and $\tau$ is the time of arrival relative to an arbitrary time base. The likelihood, or the probability of obtaining the observation $\tilde x(\omega)$ conditioned on the value of the parameter vector $\boldsymbol\theta = (\rho, \gamma, \tau)$, is Gaussian, with the log-likelihood proportional to:
%
\begin{equation}\label{eq:gaussian-likelihood}
	\mathcal{L}(\tilde x; \boldsymbol\theta) = p(\tilde x | \boldsymbol\theta)
		\propto \exp \left[
		- \frac{1}{2} \int_0^\infty \frac{\left|\tilde x (\omega)
			- \tilde h (\omega; \boldsymbol\theta) \right|^2}{S(\omega)} \, \mathrm{d}\omega
	\right].
\end{equation}

The Fisher information matrix for a measurement $\tilde x$ described by the unknown parameter vector $\boldsymbol{\theta}$ is
%
\begin{equation}\label{eq:general-fisher-matrix}
	\mathcal{I}_{jk} = \mathrm{E} \, \left[
		-\frac{\partial^2 \log
			\mathcal{L}(\tilde x ; \boldsymbol\theta)}
			{\partial \theta_j \theta_k}
	\right].
\end{equation}
%
This equation involves double derivatives, but substantial simplification is possible when---as in this case---the likelihood is Gaussian:
%
\begin{equation}\label{eq:gaussian-fisher-matrix}
	\mathcal{I}_{jk} = \int_0^\infty \Re \left[
        \left(\frac{\partial \tilde h}{\partial \theta_j}\right)^*
        \left(\frac{\partial \tilde h}{\partial \theta_k}\right)
	\right] \frac{1}{S(\omega)} \, \mathrm{d}\omega,
\end{equation}
%
where $\Re$ denotes the real part and $^*$ the complex conjugate. This form is more useful because it involves only first derivatives, and of the signal $\tilde h (\omega)$ rather than the entire observation $\tilde x (\omega)$.

If the waveform $\tilde u (\omega)$ is normalized such that $\mathcal{I}_{\rho\rho} = 1$, we can introduce the \textit{\ac{SNR}-weighted moments of angular frequency},
%
\begin{equation}\label{eq:angular-frequency-moments}
    \overline{\omega^k} =
        \left[ \int_0^\infty \frac{|\tilde u (\omega)|^2}{S(\omega)} \omega^k \, \mathrm{d}\omega \right]
        \left[ \int_0^\infty \frac{|\tilde u (\omega)|^2}{S(\omega)} \, \mathrm{d}\omega \right]
\end{equation}
%
Then, we have
%
\begin{eqnarray*}
	\mathcal{I} = \bordermatrix{
        ~ & \rho & \gamma & \tau \cr
        \rho & 1 & 0 & 0 \cr
        \gamma & 0 & \rho^2 & -\rho^2 \overline{\omega} \cr
        \tau & 0 & -\rho^2 \overline{\omega} & \rho^2 \overline{\omega^2}
    }.
\end{eqnarray*}
